Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-06-18 paper/main.tex
done with 3.4

\subsection{CAMB Data}
\label{subsec:camb}

We next investigate model fits to several cosmological 
datasets from CAMB \citep{lewis2011CAMB}. These datasets contain power spectra based
on six-dimensional cosmological parameter settings. Similar to the Mira-Titan suite, the
CAMB dataset contains data for sixty-four cosmologies, with each containing multiple 
(fifteen) low-resolution spectra, and one high-resolution spectrum. 
Unlike the Mira-Titan suite, this dataset additionally contains an infinite-resolution 
spectrum, which can be treated as the true spectrum for each cosmology,
allowing us to assess the accuracy of posterior mean estimates and effectiveness
of credible intervals. An illustration of the first cosmology from CAMB is shown
in Figure \ref{fig:fit_camb}, where the infinite-resolution spectrum is generally
much smoother than the low- and high-resolution counterparts.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{CAMB_fit_model1.png}
    \caption{Example of DGP.FCO model fit for the first batch of model runs from the CAMB data, 
             with a LOESS mean of the weighted average subtracted to emphasize details. We can 
             compare the orange lines (posterior mean and dotted 95\% credible interval) to the 
             light blue line, representing the infinite-resolution ``true" spectrum.}   
    \label{fig:fit_camb}
\end{figure}

Our goal here is to train the DGP.FCO model and accurately estimate the true spectrum for each
cosmology. To imitate the role of the perturbation theory output of Mira-Titan, we anchor 
estimates of the CAMB power spectrum at the infinite-resolution for $\log_{10}(k) < -1.7$, and 
then rely strictly on the low- and high-resolution runs for $k$ above this (note the transition
area in Figure \ref{fig:fit_camb} around the transformed $k$ value of 0.2). We first use a log-log 
regression model to obtain precision estimates $\Lambda_\ell$ and $\Lambda_h$
for the low- and high-resolution runs respectively, as outlined in Section \ref{sec:data}. 
Similarly, we set $\Lambda_c^{(ii)}=10^8$ when $\log_{10}(k_i) < -1.7$ ($0$ otherwise) for $i = 1,\dots,n$.
\textcolor{red}{(why switch from $10^5$ to $10^8$?)} 
to illustrate high confidence in the infinite-resolution run for low $k$ values.
We then take a weighted average of the CAMB runs in a similar fashion
to Mira-Titan data as described in Section \ref{sec:data}: $\bar y = \Lambda^{-1}(\Lambda_c y_c + 
\Lambda_{\ell} \bar{y}_\ell + \Lambda_h y_h)$, where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$ 
and $\bar{y}_\ell = \frac{1}{15}\sum_{r=1}^{15} y_{\ell_r}$.

With weighted average and precision information, we considered a Mat\'ern kernel for 
$\Sigma_\varepsilon$. 
\textcolor{red}{(Yes, using maximum likelihood estimation with smoothness set to $5/2$, 
and we use $\tau^2$ and $\theta$ estimates here, in the sims, and for Mira-Titan; 
these estimates are coming out of the ``opt\_matern" function we created.)}
For the CAMB data, spatial correlation decays very quickly
($\theta\approx 0$ for CAMB, compare the smoothly varying low-resolution runs of 
Figure \ref{fig:fit_camb} with the jagged low-resolution CAMB runs in Figure \ref{fig:plot_data}). 
We therefore assume no spatial dependence for CAMB and specify $\Sigma_\varepsilon = 
(\Lambda_c+\Lambda_\ell+\Lambda_h)^{-1}$.

After fitting the DGP.FCO model, an illustration of the posterior mean and its UQ can
be seen in Figure \ref{fig:fit_camb}. Notice that there are peaks where both the high-resolution
run and the weighted average are biased, but our posterior mean adjusts appropriately 
and has a more reliable estimate for the infinite-resolution run (the ``truth" in this setting). 
UQ is handled well, as precision increases with increasing $k$, which is captured in our
95\% credible intervals.

\textcolor{red}{As of now, we are not comparing to DPC/hetGP/deepgp here, so no
MSE or log scores. We do have MSEs of CAMB predictions when training on these posterior means
vs training on the infinite-resolution runs; this comes up in Figures \ref{fig:pca_preds_v_camb} 
and \ref{fig:mse_camb} later.}

\subsection{Estimating Spectra with Mira-Titan Data}
\label{subsec:mira_fit}

>>>>

\subsection{CAMB Estimation}
\label{subsec:camb}

The CAMB \citep{lewis2011CAMB} dataset contains power spectra based
on six cosmological parameter settings. It includes sixty-four cosmologies,
each containing fifteen low-resolution spectra ($y_{l_r}$) and one high-resolution spectrum ($y_h$).
Each spectrum is observed along the same grid of wavenumbers ($k_i$ for $i=1,\dots,n$). 
Unlike the Mira-Titan suite, this dataset includes an infinite-resolution 
spectrum ($y_c)$ which can be treated as the true spectrum for each cosmology,
allowing us to assess model performance. An illustration of the first cosmology 
from CAMB is shown in Figure \ref{fig:fit_camb}.  The ``true'' infinite-resolution 
spectrum is notably smoother than its low- and high-resolution counterparts.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{CAMB_fit_model1.png}
    \caption{Example of DGP.FCO model fit for the first batch of model runs from the CAMB data, 
             with a LOESS mean of the weighted average subtracted to emphasize details. We can 
             compare the orange lines (posterior mean and dotted 95\% credible interval) to the 
             light blue line, representing the infinite-resolution ``true" spectrum.
             \textcolor{red}{(@Steve - can you leave the x-axis on the usual scale?
             So -1.7 is directly visible.)}}   
    \label{fig:fit_camb}
\end{figure}

To imitate the role of the perturbation theory output of Mira-Titan, we anchor 
estimates of the CAMB power spectrum at the infinite-resolution for 
$\log_{10}(k) < -1.7$, and then rely strictly on the low- and high-resolution 
runs for $k$ above this (note the transition
in Figure \ref{fig:fit_camb} at $\log_{10}(k) = -1.7$). To compute $\bar{y}$ and $\Sigma_\epsilon$,
we first use a log-log regression model to obtain precision estimates $\Lambda_\ell$ and $\Lambda_h$
for the low- and high-resolution runs respectively, as outlined in Section \ref{sec:data}. 
To anchor estimates at low resolutions, we define the diagonal matrix $\Lambda_c$ with 
$\Lambda_c^{(ii)}=10^8$ for $\log_{10}(k_i) < -1.7$ ($0$ otherwise).
We then compute the weighted average as: $\bar y = \Lambda^{-1}(\Lambda_c y_c + 
\Lambda_{\ell} \bar{y}_\ell + \Lambda_h y_h)$, where $\Lambda = \Lambda_c + \Lambda_\ell + \Lambda_h$ 
and $\bar{y}_\ell = \frac{1}{15}\sum_{r=1}^{15} y_{\ell_r}$.
For the CAMB data, spatial correlation decays very quickly
(compare the smoothly varying low-resolution runs of Mira-Titan in Figure \ref{fig:plot_data} 
with the jagged low-resolution runs of CAMB in Figure \ref{fig:fit_camb}). 
Although we considered a Mat\'ern kernel for $\Sigma_\epsilon$ here, we found the covariance
to decay so quickly ($\theta\approx 0$) that a diagonal covariance would suffice.
Therefore, we drop spatial dependence and specify $\Sigma_\varepsilon = 
(\Lambda_c+\Lambda_\ell+\Lambda_h)^{-1}$.

We fit our DGP.FCO model to the low- and high-resolution runs of each CAMB cosmology for
later use in Section \ref{sec:pred}.  For now, we provide an illustration of the DGP.FCO
prediction (solid yellow) and credible intervals (dashed yellow) for a single 
CAMB cosmology in Figure \ref{fig:fit_camb}.  Our posterior mean provides an effective estimate
of the ``true'' infinite-resolution run.  It aptly smooths over the peaks where 
both the high-resolution run and the weighted average significantly deviate from the truth
($\log_{10}(k)\approx 0.4,0.7$ - \textcolor{red}{@Steve - please adjust these numbers after 
changing the x-axis of the figure}). Our model also provides effective UQ, appropriately shrinking
credible intervals as precision increases for larger $k$.

\subsection{Mira-Titan Estimation}
\label{subsec:mira_fit}
